---
title: "Supplementary materials"
format:
  pdf:
    documentclass: scrartcl
    number-sections: true
    number-depth: 3
    #keep-tex: true
bibliography: suppmat.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
options(knitr.kable.NA = "**")
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(ggpmisc))


mp_models_data <- readRDS(
  "data/proc/monkeypox-selected-countries-model-data.rds"
)

mp_models_results <- readRDS(
  "data/proc/monkeypox-selected-countries-model-results.rds"
)

mp_selected_countries <- mp_models_data %>% 
  group_by(Country) %>% 
  summarise(
    Earliest = min(Date_confirmation),
    Latest = max(Date_confirmation)
  ) %>% 
  left_join(
    mp_models_results %>% 
      filter(term == "days") %>% 
      select(
        Country,
        nobs
      ),
    by = "Country"
  )
```

\appendix

\renewcommand{\thesection}{\Alph{section}}

## Regression models of exponential growth {#unnumbered}


### General equations

During the first part of the development of an epidemic, we can assume an exponential growth according to @eq-expeq

$$
y = \alpha \beta^{\gamma x} = \alpha e^{(\gamma \log{\beta}) x}
$$ {#eq-expeq}

where: $y$ = is the cumulative case count, $x$ = number of days since the start of the series

This equation can be linearized using logarithms, shown in @eq-logeq, and its equivalent @eq-altlogeq

$$
\log{y} = \log{\alpha} + (\gamma \log{\beta}) x
$$ {#eq-logeq}

$$
\log{y} = A + B x
$$ {#eq-altlogeq}

where: $A = \log{\alpha}$, $B = \gamma \log{\beta}$

### Model data selection

The MPX data obtained from the Global Health Data Science Initiative, was filtered using the following procedure:

1. Only cases with a confirmed status were kept
2. Data for which the confirmation date ranged from the epidemiological weeks 20 to 33 of 2022 were used. This allowed us to consider only cases in non-endemic countries
3. Data was combined by confirmation date at the country level, and, after ordering each country timeseries, a cumulative number of cases was calculated
4. Data from countries with 200 or more cumulative cases, up to the mos recent reported date, were considered for modeling
5. From the selected countries, we used only dates for which the number of cumulative cases was equal or greater than 10.
6. For each country timeseries, we created a days count series which reflected the difference between the earliest date in the series and the current one.

At the end of this selection procedure, we obtained the list in the following table:

```{r}
kbl(
  mp_selected_countries,
  format = "latex",
  booktabs = TRUE,
  caption = "Countries selected for modeling",
  col.names = c(
    "Country",
    "Earliest",
    "Latest",
    paste0("N° Obs.", footnote_marker_number(1, "latex"))
  ),
  align = "lccr",
  escape = FALSE
) %>% 
  kable_styling(
    latex_options = c("striped", "hold_position")
  ) %>% 
  add_header_above(
    c(
      " " = 1,
      "Confirmation dates" = 2,
      " " = 1
    )
  ) %>% 
  column_spec(1, italic = TRUE) %>% 
  footnote(
    number = "Number of days with reports of confirmed cases in the date range",
    threeparttable = TRUE
  )
```
 
### Regression results

Using @eq-altlogeq, we performed a regression of the logarithm (base 10) of the cumulative number of cases versus the days counts (*vide supra*)

The results of the regressions can be seen in the following table

```{r}
tbl_df <- mp_models_results %>% 
  mutate(
    term = if_else(
      term == "days",
      "B",
      "A"
    )
  ) %>% 
  select(
    Country,
    term,
    estimate,
    std.error
  ) %>% 
  pivot_wider(
    names_from = "term",
    values_from = c(estimate, std.error)
  ) %>% 
  left_join(
    mp_models_results %>% 
      filter(term == "days") %>% 
      select(
        Country,
        adj.r.squared,
        p.value,
        nobs
      ),
    by = "Country"
  ) %>% 
  relocate(
    4,
    .after = 2
  ) %>% 
  mutate(
    pval = if_else(p.value < 0.001, "< 0.001", as.character(p.value))
  ) %>% 
  select(-p.value) %>% 
  relocate(
    pval,
    .before = "nobs"
  ) 

kbl(
  tbl_df,
  digits = 4,
  col.names = c(
    "Country",
    "Intercept",
    "$\\text{S.E.}_{inter.}$",
    "Slope",
    "$\\text{S.E.}_{slope}$",
    "$R^2_{adj}$",
    "p-value",
    paste0("N° Obs.", footnote_marker_number(1, "latex"))
  ),
  format = "latex",
  align = "lccccccr",
  booktabs = TRUE,
  escape = FALSE,
  caption = "Regression results for the selected countries"
) %>% 
  kable_styling(
    latex_options = c("striped", "hold_position")
  ) %>% 
  add_header_above(
    c(
      " " = 1,
      "Parameters" = 4,
      "Statistics" = 3
    )
  ) %>% 
  column_spec(1, italic = TRUE) %>% 
  footnote(
    number = "Number of days with reports of confirmed cases in the date range",
    threeparttable = TRUE
  )
```

### Plots of the regression results per country

```{r}
#| fig.width: 7
#| fig.height: 8
#| eval: true
mp_results <- readRDS("data/proc/monkeypox-selected-countries-model-results.rds") %>%
  filter(term == "days") %>% 
  select(
    Country,
    adj.r.squared,
    nobs,
    p.value
  )
mp_mod_data <- readRDS("data/proc/monkeypox-selected-countries-model-data.rds") %>%
  left_join(
    mp_results,
    by = "Country"
  )

countries <- as.character(unique(mp_mod_data$Country))

mk_reg_plot <- function(df) {
  country <- unique(df$Country)
  nobs <- unique(df$nobs)
  plt <- ggplot(df, aes(x = days, y = log(acc_cases))) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x) +
    stat_regline_equation(
      #label.y.npc = 0.95,
      formula = y ~ x,
      size = 5,
      aes(
        label = paste(
          ..eq.label.., 
          ..adj.rr.label..,
          sep = "~`,`~"
        )
      )
    ) +
    annotate(
      geom = "text_npc",
      npcx = 0.8,
      npcy = 0.15,
      label = glue::glue("N = {nobs}"),
      hjust = 0,
      size = 5
    ) +
    stat_cor(
      aes(label = paste(..p.label..)),
      p.accuracy = 0.001,
      label.y.npc = 0.1,
      label.x.npc = 0.8,
      hjust = 0,
      size = 5
    ) +
    labs(
      x = "Number of days",
      y = "log(Cumulative cases)",
      title = country
    ) +
    theme_bw(base_size = 16) +
    theme(
      plot.title = element_text(size = 24, face = "bold")
    )
  return(plt)
}

for (cntry in countries) {
  df <- mp_mod_data %>% 
    filter(Country == cntry)
  print(mk_reg_plot(df))
}

```


## Estimation of the duplication time

Given two cumulative counts: $C_1$, $C_2$, obtained at two different times: $t_1$, $t_2$, their ratio $Q$ is defined as follows:

$$
Q = \frac{C_2}{C_1} = \frac{\alpha e^{(\gamma \log{\beta}) t_2}}{\alpha e^{(\gamma \log{\beta}) t_1}} = \alpha e^{(\gamma \log{\beta}) (t_2 - t_1)}
$$ {#eq-rdef}

When $Q = 2$, then the difference ($t_2 - t_1$) is defined as the duplication time $T_d$: the time it takes for the count to double in size with respect to an initial value.

Using that definition, and employing @eq-altlogeq:

\begin{align}
\log{Q} &= \log{\frac{C_2}{C_1}} \\
  &= \log{C_2} - \log{C_1} \\
  &= (A + B t_2) - (A - B t_1) \\
  &= B(t_2 - t_1) \\
  &= B(T_d) 
\end{align}

Finally we can define $T_d$ in terms of the slope ($B$), if $Q = 2$, then:

$$
T_d = \frac{\log{2}}{B}
$$ {#eq-td}

Using the regression results of the slope and its standard error, we can calculate $T_d$ and its C.I., obtaining:

```{r}
tbl_td <- tbl_df %>% 
  select(
    Country,
    nobs,
    B = estimate_B,
    SEB = std.error_B
  ) %>% 
  mutate(
    low_b = B - 1.96 * SEB,
    hi_b = B + 1.96 * SEB,
    CI_B = glue::glue(
      "[{sprintf('%.4f', low_b)} - ",
      "{sprintf('%.4f', hi_b)}]"
    ),
    td = log(2) / B,
    low_td = log(2) / hi_b,
    hi_td = log(2) / low_b,
    CI_td = glue::glue(
      "[{sprintf('%.4f', low_td)} - ",
      "{sprintf('%.4f', hi_td)}]"
    )
  ) %>% 
  select(
    Country,
    nobs,
    B,
    SEB,
    CI_B,
    td,
    CI_td
  ) %>% 
  arrange(td)

kbl(
  tbl_td,
  digits = 4,
  col.names = c(
    "Country",
    paste0("N° Obs.", footnote_marker_number(1, "latex")),
    "Slope",
    "$\\text{S.E.}_{slope}$",
    "$\\text{95\\% CI}_{slope}$",
    "$\\widehat{T}_d$",
    "$\\text{95\\% CI}_{\\widehat{T}_d}$"
  ),
  format = "latex",
  booktabs = TRUE,
  align = "lcccccc",
  linesep = "",
  caption = "Duplication time estimates for selected countries",
  escape = FALSE
) %>% 
  kable_styling(
    latex_options = c("striped", "HOLD_POSITION",
                      "scale_down")
  ) %>% 
  column_spec(1, italic = TRUE) %>% 
  column_spec(6, border_left = TRUE) %>% 
  footnote(
    number = "Number of days with reports of confirmed cases in the date range",
    threeparttable = TRUE
  )
```

\newpage

## Estimation of the effective reproduction number

According to [@bonifaziSimplifiedEstimateEffective2021] there is a functional relation between the effective reproduction number ($R_t$), and the duplication time, of the form:

$$
\widehat{R}_t = e^{(g \log{2}) / \widehat{T}_d}
$$ {#eq-rt-td}

where: $g$ is the generation time, and $\widehat{T}_d$ is the estimate of the duplication time.

Combining @eq-td and @eq-rt-td, we can derive

$$
\widehat{R}_t = e^{(g \log{2}) / (\log{2} / B)} = e^{g B}
$$ {#eq-rt-b}

A recent article [@guzzettaEarlyEstimatesMonkeypox2022], gives an early estimate for the generation time of the current MPXV outbreak of 12.5 days (95% CI: [7.5 - 17.3]). Using @eq-rt-b, along with our estimate for the slope ($B$) and the published estimate for $g$ with its 95% CI, we can compute a possible value of $R_t$:


```{r}
tbl_rt <- tbl_df %>% 
  select(
    Country,
    B = estimate_B
  ) %>% 
  mutate(
    rt = exp(12.5 * B),
    lo_rt = exp(7.5 * B),
    hi_rt = exp(17.3 * B)
  ) %>% 
  arrange(desc(rt))

kbl(
  tbl_rt, 
  booktabs = TRUE, 
  digits = 3,
  format = "latex",
  col.names = c(
    "Country",
    "Slope",
    paste0("$R_t$(mean)", footnote_marker_number(1, "latex")),
    paste0("$R_t$(lower)", footnote_marker_number(2, "latex")),
    paste0("$R_t$(upper)", footnote_marker_number(3, "latex"))
  ),
  escape = FALSE, 
  linesep = "",
  caption = "Estimates of $R_t$ for selected countries"
) %>% 
  kable_styling(
    latex_options = c("striped", "HOLD_POSITION",
                      "scale_down")
  ) %>% 
  column_spec(1, italic = TRUE) %>% 
  footnote(
    number = c(
      "Using the mean estimate of 12.5 days",
      "Using the lower estimate of 7.5 days",
      "Using the higher estimate of 17.3 days"
    )
  )

```


\newpage

```{r}

# generate the incidence table
mk_incidence_tbl <- function(df, row_indexes, last_epiweek, caption) {
  df <- df %>% 
    mutate(
      ini_date = if_else(
        ini_date >= as.Date("2022-05-01"),
        as.character(ini_date),
        paste0(ini_date, footnote_marker_number(2, "latex"))
      )
    )
  
  kbl(
    df,
    booktabs = TRUE,
    longtable = TRUE,
    digits = 3,
    format.args = list(
      big.mark = ",",
      scientific = FALSE
    ),
    col.names = linebreak(
      c(
        "Country",
        paste0("Date of first\nconfirmed case", 
               footnote_marker_number(1, "latex"),
               footnote_marker_number(3, "latex")
               ),
        paste0("Total\ncases", 
               footnote_marker_number(1, "latex"),
               footnote_marker_number(3, "latex")),
        paste0("Population\n(2022)",
               footnote_marker_number(4, "latex")),
        paste0("Incidence\n(per million)", 
               footnote_marker_number(3, "latex"))
      )
    ),
    align = "lcrrr",
    caption = caption, 
    escape = FALSE
  ) %>% 
    pack_rows(
      index = row_indexes,
      latex_gap_space = "1em",
      underline = TRUE
    ) %>%
    kable_styling(
      latex_options = c("striped", "repeat_header", 
                        "hold_position")
    ) %>% 
    row_spec(0, bold = TRUE) %>% 
    footnote(
      number = c(
        "Source: Global.health Monkeypox data repository",
        "Reports earlier than May 2022 are from endemic areas",
        paste0("As of complete epidemiological week #", last_epiweek, " of 2022."),
        "Source: UN 2022 Revision of World Population Prospects"
      )
    ) %>% 
    column_spec(1, italic = TRUE)
}

# read the data
mp_df <- readRDS("data/proc/monkeypox-confirmed-augmented.rds") %>%
  filter(complete_epiweek)
last_epiweek <- max(mp_df$epiweek, na.rm = TRUE)

tbl_df <- mp_df %>%
  group_by(
    Country,
    Continent,
    Region,
    PopTotal,
    PopDensity
  ) %>%
  summarise(
    ini_date = min(Date_confirmation, na.rm = TRUE),
    n_cases = n()
  ) %>%
  mutate(
    incidence = 1e6 * n_cases / PopTotal
  ) %>%
  arrange(Continent, Region, desc(incidence)) %>%
  ungroup()

incid_tbl_df <- tbl_df %>% 
    ungroup() %>% 
    select(
      Continent,
      Country,
      ini_date,
      n_cases,
      PopTotal,
      incidence
    ) %>% 
    arrange(
      Continent,
      desc(incidence)
    ) %>%
  ungroup()
```


## Cumulative incidence for all countries with confirmed cases

In the following two tables, we show the cumulative incidence up to the most current complete epidemiological week (2022W`r last_epiweek`), separating those countries with 10 or more confirmed cases in total, from those with less than 10 cases.

```{r}
ten_or_more <- incid_tbl_df %>% 
  filter(n_cases >= 10)

row_indexes <- ten_or_more %>% 
  group_by(Continent) %>% 
  summarise(
    max_row = max(row_number())
  ) %>% 
  deframe()

mk_incidence_tbl(
  ten_or_more %>% select(-Continent), 
  row_indexes, 
  last_epiweek,
  "Cumulative incidence per million for countries at least 10 confirmed cases"
)


```


```{r}
less_than_then <- incid_tbl_df %>% 
  filter(n_cases < 10)

row_indexes <- less_than_then %>% 
  group_by(Continent) %>% 
  summarise(
    max_row = max(row_number())
  ) %>% 
  deframe()

mk_incidence_tbl(
  less_than_then %>% select(-Continent), 
  row_indexes, 
  last_epiweek,
  "Cumulative incidence per million for countries with less than 10 confirmed cases"
)

```

## References

::: {#refs}
:::
